 \mylabel{sec-ci}
% \begin{minipage}[b]{0.5\textwidth}
 The following tables can be used to determine confidence
 intervals for quantiles (including median),
 according to
 \thref{theo-conf-ci-median}.
%\end{minipage}
%
%\hfill HERE COMES A DRAWING\\
For a sample of $n$ iid data points
$x_1,...,x_n$, the tables give a confidence
interval at the confidence level $\gamma=0.95$ or
$0.99$ for the $q$-quantile with $q=0.5$
(median), $q=0.75$ (quartile) and $q=0.95$. The
confidence interval is $[x_{(j)},x_{(k)}]$, where
$x_{(j)}$ is the $j$th data point sorted in
increasing order.

The confidence intervals for $q=0.05$ and $q=0.25$ are not given
in the tables. They can be deduced by the following rule. Let
$[x_{(j)},x_{(k)}]$ be the confidence interval for the
$q$-quantile given by the table.  A confidence interval for the
$1-q$-quantile is $[x_{(j')},x_{(k')}]$ with
 \bearn
 j' = n+1 -k \\
 k' = n+1 -j
 \eearn
 For example, with $n=50$, a confidence interval for the third quartile
 ($q=0.75$) at confidence level $0.99$ is $[x_{(29)},x_{(45)}]$,
 thus a confidence interval for the first quartile ($q=0.25$) at
 confidence level $0.99$ is $[x_{(6)},x_{(22)}]$.

For small values of $n$ no confidence interval is
possible. For large $n$, an approximate value is
given, based on a normal approximation of the
binomial distribution.

\begin{petiteNote} The tables give $p$, the actual confidence
level obtained, as it is not possible to obtain a
confidence interval at exactly the required
confidence levels. For example, for $n=10$ and
$\gamma=0.95$ the confidence interval given by
the table is $ \left[X_{(2)},X_{(9)}\right]$; the
table says that it is in fact a confidence
interval at level $0.979$.

The values of $j$ and $k$ are chosen such that
$j$ and $k$ are as symmetric as possible around
$\frac{n+1}{2}$. For example, for $n=31$ the
table gives the interval
$\left[X_{(10)},X_{(22)}\right]$. Note that this
is not the only interval that can be obtained
from the theorem. Indeed, we have:\begin{quote}
\begin{tabular}{c c c}
  % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
  $j$ & $k$ &  $\P\left(X_{(j)}< m_{0.5} < X_{(k)}\right)$\\
  \hline
  \input{bin31ci.dat}%
\end{tabular}%
  \nfs{binomial.m}%
\end{quote}
Thus we have \imp{several} possible confidence
intervals. The table simply picked one for which
the indices are closest to
 being symmetrical around the estimated median, i.e. the indices $j$ and $k$ are equally
 spaced around $\frac{n+1}{2}$, which is used for estimating the median. In some cases,
  like $n=32$, we do not find such an interval exactly; we have for instance:
  \begin{quote}
\begin{tabular}{c c c}
  % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
  $j$ & $k$ &  $\P\left(X_{(j)}< m_{0.5} < X_{(k)}\right)$\\
  \hline
  \input{bin32ci.dat}
\end{tabular}%
  \nfs{binomial.m}%
\end{quote}
Here, the table arbitrarily picked the former.
\end{petiteNote}


\input{tablesCI}
%\center \footnotesize
%\begin{tabular}{|c||c|c|c|}\hline
% $n$ & $j$ & $k$ & $p$ \\ \hline \hline
%\input{binlist95.dat}
% \hline
%$n \geq 50$ & $\frac{n}{2}- 0.980\sqrt{n}$ &$\frac{n}{2}+1+
%0.980\sqrt{n}$ & 0.95 \\ \hline
%\end{tabular}
%\hspace{0.0cm}
%\begin{tabular}{|c|c|c|}\hline
%$j$ & $k$ & $p$ \\ \hline  \hline
%\input{binlist99.dat}
% \hline
% $\frac{n}{2}- 1.288\sqrt{n}$ &$\frac{n}{2}+1+ 1.288\sqrt{n}$ &
%0.99 \\ \hline
%
%\end{tabular}\mycaption{Confidence Interval
%for the median at confidence levels at least $0.95$ (left) and
%$0.99$ (right). For a sample of $n$ iid data points $x_1,...,x_n$,
%a confidence interval for the median is $[x_{(j)},x_{(k)}]$, where
%$x_{(j)}$ is the $j$th data point in increasing order: $p$ is the
%actual confidence level obtained (it is not possible to obtain a
%confidence interval at exactly the required confidence levels).
%Cells marked ``n/a" mean that no confidence interval is possible
%($n$ is too small).}
% \mylabel{tab-conf-medianci}
%\end{table}Table of confidence intervals for median:
%\nfs{Tables completes
%\begin{table}
%\begin{tabular}{|c|c|c|c||c|c|c|c|c|c|c|}
%\hline  $n$ & $j$ & $k$ & $p$ & $j$ & $k$ & $p$& $j$ & $k$ & $p$ & Best\\
%\hline \hline
%\input{binlist95-tout.dat}
%\hline
%
%\end{tabular} \mycaption{0.95}
%\end{table}
%\begin{table}
%\begin{tabular}{|c|c|c|c||c|c|c|c|c|c|c|}
%\hline  $n$ & $j$ & $k$ & $p$ & $j$ & $k$ & $p$& $j$ & $k$ & $p$ & Best\\
%\input{binlist99-tout.dat}
%\hline
%
%\end{tabular} \mycaption{0.99}
%\end{table}
%}
